In the Age of Web:\\ \textnormal{\LARGE Typed Functional-First Programming Revisited}

### 2015-10-25 ml-2014-paper/paper.tex
Camera ready version of ML2014 post-proceedings

\author{Tomas Petricek
\institute{University of Cambridge, UK}
\email{tomas@tomasp.net}
\and
Don Syme
\institute{Microsoft Research Cambridge, UK}
\email{don.syme@microsoft.com}
}
\def\titlerunning{In the Age of Web: Typed Functional-First Programming Revisited}
\def\authorrunning{T. Petricek, D. Syme}
\begin{document}
\maketitle

@@

Perhaps the most remarkable assumption is the idea that programs fundamentally operate in a closed
world. Although we have learned how to perform FFI and I/O \cite{haskell-ffi,haskell-imperative},
those are treated as dealing with the ``dirty real world''. For a practical solution, we argue that
we need to go much further, but deeper integration with (untyped) JavaScript libraries and (evolving)
services inevitably breaks strict type safety requirements.

@@

Note that we use a standard F\# function \ident{List.map} to iterate over the countries. This has
a side-effect of creating the HTML elements, but it also returns a new list. The result is a list
of $\ident{string}\ast\ident{Indicators}\ast\ident{jQuery}$ values representing the country name,
its indicators (for accessing the World Bank data) and the created DOM object representing the
checkbox.

@@

The aim of this paper is not to make value judgements and argue what is better. Using the case
study as a basis, we claim that the outlined approach is just \emph{one possible} and that it
\emph{works in practice}. In the rest of the paper, we give more details about the most important
aspects of the approach and discuss alternatives.

@@

In technical terms, type providers are libraries that are loaded by the compiler (and editor)
and are executed at compile-time (or edit-time). A type provider builds information about types
it provides and makes those available to the compiler. It is worth noting that this is done lazily,
so the type provider does not need to provide types for the entire information space at once.

In terms of programming language theory, type providers can be easily explained anecdotally.
They change the starting point for a type-checking of a program as follows:
%
\begin{equation*}
\hspace{8em}
\begin{array}{rclll}
 \emptyset &\narrow{\vdash}& e : \tau
    &\qquad& (\textit{classical, or closed-world})\\
 \pi~(\includegraphics[width=1em,trim=0 1.5em 0 0]{earth.png})  &\narrow{\vdash}& e : \tau
    &\qquad& (\textit{type providers, or open-world})
\end{array}
\end{equation*}
%
Rather than starting the type-checking with an empty context, we start type-checking with a
context containing a \emph{projection} of some information from the world. In this paper, we do not
attempt to formalize type providers, so we avoid the question of what, formally, is the world here.
However, this analogy is useful for understanding how type providers work and how they can go wrong:

\begin{itemize}
\item \emph{Type provider failure.}
  The projection (a type provider) is implemented in F\# and can fail (for example, if the internet
  is required but unavailable). However, when it succeeds, it generates valid F\# code and so
  unchecked compilation errors cannot happen.

@@

\item \emph{Recompilation failure.}
  When the world changes and we \emph{recompile} the program, the result of the projection can
  differ and so we may not be able to type-check code that used to type-check. This can be seen as
  negative aspect, but there is a clear benefit -- it means we discover an error that would happen
  at runtime earlier, that is during recompilation.
\end{itemize}

\noindent
The above points are \emph{general} points about type providers. In this paper, we are more interested
in the \emph{specific} type providers used in the case study. We look at the World Bank and TypeScript
type providers in the following two sections. When discussing type providers, we need to look at
how the provider (or the projection $\pi$) works and what assumptions it makes about the world.
That is, what world changes cause the provided type to change and what world changes cause a
runtime exception.

@@

The World Bank provider \cite{fsharp-data} is a type provider for a specific data source -- it has
been designed specifically for World Bank data. This contrasts with F\# Data type providers for
working with CSV, XML and JSON data that take a sample or schema as a parameter and can thus be used
with files of any structure.

\subsubsection{The World Bank type provider}

The projection implemented by the type provider generates a type called \ident{Countries} that contains
countries (and regions) as members. Each member returns a value of type \ident{Indicators} that is
also generated and contains all indicators as members. Note that the type provider generates just
these two types -- in particular, it does not generate a new type for each country. This would be an
interesting option if we could check what indicators are available for a given country.

The type provider builds the structure of the provided types, but the provided types are \emph{erased}
during compilation and replaced with runtime implementation. For the World Bank type provider, the
erasure works as follows:
%
\begin{equation*}
\begin{array}{lcl}
 \sem{~\ident{data.}\lident{Czech Republic}~} &=&
   \ident{data.GetCountries}().\ident{GetCountry}(\str{CZE}) \\
 \sem{~\ident{cz.}\lident{School enrollment, tertiary (\% gross)}~} &=&
   \ident{cz.AsyncGetIndicator}(\str{SE.TER.ENRR})
\end{array}
\end{equation*}
%
The underlying operations (\ident{GetCountry}, \ident{GetCountries} and \ident{AsyncGetIndicator})
are operations of an underlying runtime library (which is a normal non-provided library). This is
a typical pattern -- type providers generally produce a light layer on top of a rich runtime library.

An important point here is that the type provider uses the \emph{name} of a country or an indicator
for the member name, but at runtime, it uses a \emph{code} of the country or the indicator. This has
interesting implications for the safety properties, as discussed in the next section.

@@

Now that we know how the World Bank type provider generates the types and the runtime code, we can
look at the assumptions it makes about the world and in which cases it encounters one of the failures
defined in Section~\ref{sec:tp-def}.

\begin{itemize}
\item If you are offline when using the type provider (compiling or editing), the type provider will
  not be able to obtain the list of countries or indicators. This is an example of \emph{type provider
  failure}. In case of the World Bank, this is partly prevented by caching of the schema, but internet
  access is required when using the provider for the first time.

\item An interesting case is when a country or indicator is renamed. In that case, compiled code
  using the World Bank type provider will continue to work (there is no \emph{runtime failure}),
  because the lookup uses the country or the indicator code. However, recompilation will fail
  as the member name (which appears in the source code) will be different.

\item Finally, if a country or indicator completely disappears, we will get \emph{runtime failure}
  when running existing compiled code and \emph{recompilation failure}. here, the \emph{recompilation
  failure} is a useful indicator that the assumptions made by our code have been violated by
  the data source.
\end{itemize}

\noindent
When contrasted with the traditional view of safety in the ML family of languages, the World Bank
type provider certainly relaxes the usual safety conditions. However, this is not the case when
compared with standard developer practices. In particular, if we wrote
$\ident{data.GetCountries}().\ident{GetCountry}(\str{CZE})$ by hand, the \emph{runtime failure}
behaviour would be the same, but we would lose the useful \emph{recompilation failure}. The following
section discusses this (and other) alternatives.

@@

There are two main alternatives that are worth discussing. First, how would the code look if
we did not use any type provider. Second, what are alternative designs for the World Bank type
provider and how would such alternatives change the safety properties.

\vspace{-1em}
\paragraph{Accessing countries without type providers.} First, consider how we would write the code to obtain a list
of countries without using the World Bank type provider. Assuming we want only the countries and
regions in Figure~\ref{fig:wb}, we could specify the country codes and then obtain the country objects:
%
\begin{equation*}
\begin{array}{l}
 \kvd{let}~\ident{data}~=~\ident{WorldBank.GetDataContext}() \\
 \kvd{let}~\ident{countryCodes}~=~\lbrack~\str{EU},~\str{CZE},~\str{GBP},~\str{USA}~\rbrack \\
 \kvd{let}~\ident{countries}~=~\ident{countryCodes}~|\rangl~\ident{List.map}~(\ident{data.GetCountries}().\ident{GetCountry})
\end{array}
\end{equation*}
%
The motivation for the case study was a data journalism application with a transparent logic
that is easy to understand (and perhaps even modify). The type provider certainly achieves these
goals better -- a country can be added just by looking through an auto-complete list and we do not
need to know its code or exact name. However, if we wanted to list all countries in a region, then
the World Bank type provider would not be helpful.

As for the safety properties, as discussed earlier, using explicit country codes has the same
runtime properties as using the type provider. This could be avoided by listing all countries in
a given region (or, according to another rule), but that is solving a different problem.

In summary, the World Bank type provider fits a specific niche that we were exploring in the case
study. That is, when we want to create an information analysis that accesses specific data from
an information source. With rising popularity of data journalism and Open Government Data
initiatives, we argue that this is an important problem domain. Here, we first specify the country
codes and then obtain the country objects.

@@

However, that is only shifting the burden of error handling from the library to the user. If the
author of the application does not intend to handle errors (for example, by skipping missing data),
they would have to write the following:
%
\begin{equation*}
\begin{array}{l}
 \kvd{let}~\ident{optCountries}~= \\
 \quad~\lbrack~~ \ident{data.Countries.}\lident{European Union} \\
 \qquad   \ident{data.Countries.}\lident{Czech Republic} \\
 \qquad   \ident{data.Countries.}\lident{United Kingdom} \\
 \qquad   \ident{data.Countries.}\lident{United States} ~\rbrack
\\[0.5em]
 \kvd{let}~\ident{countries}~=\\
  \quad \ident{optCountries}~|\rangl~\ident{List.map}~(\kvd{fun}~\ident{countryOpt}~\rightarrow \\
  \qquad \kvd{match}~\ident{countryOpt}~\kvd{with}~\ident{Some~c}~\rightarrow~\ident{c}~|~\ignp~\rightarrow
    \ident{failwith}~\str{Country does not exist!})
\end{array}
\end{equation*}
%
For the purpose of our case study, providing option types would only make the code more complex.
However, if we wanted to handle errors more gracefully, then option types would be a better
alternative. Note, however, that we can still handle exceptions using $\kvd{try}~\ldots~\kvd{with}$.

Whether using option types is more desirable or not often depends on the use case (how we
want to handle errors) and reliability of the data source (countries do not disappear often).
This alternative can be supported by adding a static parameter to the type provider (like
$\ident{Asynchronous}=\kvd{true}$). For example, the CSV type provider from F\# Data has a
parameter \ident{AssumeMissingValues} that instructs the type provider to provide option values,
even if the sample does not contain them.

\vspace{-1em}
\paragraph{Accessing data ``as of time''.}
So far, we discussed how to mitigate the problems caused by the fact that the open world changes.
Can we make the world \emph{not} change? This is a sensible question for some data sources
(although not for the World Bank). For example Freebase \cite{google-freebase} supports the
notion of ``as of time''. When calling Freebase with the \ident{as\_of\_time} parameter, the API
returns data and meta-data that was present at the specified date.

This would be a desirable option for our case study -- if we want to illustrate point made
by an article, the accompanying application could use data available at the time when the
article was written. However, this feature relies on the ability of the data source (not all
services have this), or on the possibility to create local snapshots (difficult for large data
sources).

@@

\subsubsection{The TypeScript type provider}
The projection implemented by the TypeScript type provider imports interface definitions in the
\str{.d.ts} format specified by TypeScript \cite{ms-typescript}. This is a format of type
annotations for JavaScript libraries. The following example is an excerpt from the \str{jquery.d.ts}
file that describes the \ident{jQuery} value and the \ident{attr} method used in our example:
%
\begin{equation*}
\begin{array}{l}
 \kvd{declare var}~\ident{jQuery}~:~\ident{JQueryStatic};
\\[0.5em]
 \kvd{interface}~\ident{JQueryStatic}~\{ \\
 \quad (\ident{selector}: \kvd{string}, \ident{context?}: \kvd{any}):~\ident{JQuery}; \\
\}
\\[0.5em]
\kvd{interface}~\ident{JQuery}~\{ \\
\quad \ident{attr}(\ident{attributeName}~:~\kvd{string}):~\kvd{string}; \\
\quad \ident{attr}(\ident{attributeName}~:~\kvd{string},~\ident{value}~:~\kvd{any}):~\ident{JQuery}; \\
\}
\end{array}
\end{equation*}
%
The example demonstrates a number of typical problems that arise when using type providers for
language interoperation. Not all TypeScript language constructs have direct equivalent in F\# and
the TypeScript type provider needs to map them to other constructs that are available:
%
\begin{itemize}
\item The interface file defines \ident{jQuery} as a global variable. Although F\# supports
  global bindings, those cannot be provided by a type provider and so the global variable
  is exposed as a static member of the imported type, so we use it via \ident{j.jQuery}.

@@

\noindent
It is also worth noting that the \ident{attr} member is overloaded and uses an optional parameter.
However, both of these features are available in F\# and can be directly mapped. Now that we
covered how the type structure is generated, we also need to show what code is generated when
the types are erased:
%
\begin{equation*}
\begin{array}{l}
 \sem{~\ident{j.jQuery.Invoke}(\ident{command})~} ~= \\
 \qquad \kvd{let}~\ident{jq}~=~\ident{Emit.PropertyGetImpl}(\kvd{true}, \str{jQuery}, \lbrack|~|\rbrack) \\
 \qquad \ident{Emit.CallImpl}(\kvd{false}, \str{}, \lbrack|~ \ident{jq};~ \ident{command}~|\rbrack~) \\[0.75em]
 \sem{~\ident{jQuery}(\str{<input>})\ident{.attr}(\str{type},\str{checkbox})~} ~= \\
 \qquad \ident{Emit.CallImpl}(\kvd{false}, \str{attr}, \lbrack|~ \ident{jQuery}(\str{<input>});~\str{type};~\str{checkbox}~|\rbrack~)
\end{array}
\end{equation*}
%
An important point here is that the above provided code is never actually \emph{executed}.
As discussed in Section~\ref{sec:js-meta}, it is translated to JavaScript (which then runs
in the web browser). So, the above code can be seen more as instructions for the translator.
The mapping is quite simple -- member calls are translated to \ident{Emit.CallImpl} (empty
name denotes that the object itself is called) and property getters are translated to
\ident{Emit.PropertyGetImpl}. In both cases the first argument denotes whether the call is
static and the last argument is an array of arguments.

\subsubsection{Safety of cross-language type providers}
As mentioned earlier, TypeScript does not have the type soundness property \cite{ms-safets},
because of covariant generics. This illustrates two points. First, this is a property of
\emph{running} TypeScript code, which is not the case here -- we use TypeScript only as a
source of annotations for JavaScript libraries. There is no guarantee that the library will
actually adhere to the specification, which is an issue that we return to when discussing
translation to JavaScript (Section~\ref{sec:js-meta}).

The second point is that we can not import the typing rules associated with TypeScript.
The type provider mechanism can only provide F\# types that then behave according to the
F\# typing rules. This means that we can import generics\footnote{Due to technical
limitations, this is not currently allowed by F\# type providers, but there are no
fundamental reasons for this.}, but this will not automatically allow using them
covariantly. In some cases, the F\# type system is simply more strict -- and in that case,
users have to use unsafe operations, such as $\ident{unbox}\langl \ident{bool} \rangl$
in the \ident{render} function. We return to the topic of unsafe operations in
Section~\ref{sec:js-meta}.

The above discussion highlights a broader point about using type providers for language
interoperability. Given any language, its type system be weaker or stricter in some ways.
Both cases make mapping difficult:

\begin{itemize}
\item First, consider a weaker system with less type information. The type provider may need
  to map more types to \ident{object} (or similar general type). However, this will make
  the provided operations hard to use (because the source language is more flexible than
  F\#). An alternative is to require explicit annotations (like \textcolor{strclr}{\ttfamily d.ts}
  files for JavaScript).

\item Second, if the imported language has a more precise type system than F\#, the type
  provider has to drop some of the information. This can be done safely for types in
  contravariant positions. In the unsafe case, the runtime needs to perform dynamic checks
  (Section~\ref{sec:conc-interop}).
\end{itemize}

\noindent
In case of JavaScript, we encounter both cases. In many cases, the return type is not
statically known and is exposed as \kvd{any}, which we then map to F\# \ident{object}
(and the developer has to use \ident{unbox}). However, TypeScript also supports limited
form of dependent typing (overloading on constants \cite{ms-typescript-09}). This cannot
be expressed in the F\# type system and the developer has to do more work (for example,
choose the right overload or provide an explicit type annotation).

An interesting example of type provider for a language with a weaker type system is the
R provider \cite{fsharp-rprovider}. It solves the mismatch by using runtime reflection
to discover R functions and their declared parameters. However, R provider also generates
additional function overloads that let the user specify additional parameters (not explicitly
declared in the signature). To make such interoperability easier, we proposed additional
escape mechanism \cite{fsharp-paramsdict} that provide usable syntax when static type information
cannot be fully recovered.

@@

\vspace{-1em}
\paragraph{Embedding inline JavaScript.}
To make the interoperability mechanism general, the host language (here, F\#) should not know about
the syntax of the embedded language (here, JavaScript). This means that the JavaScript code can
only be embedded as strings. FunScript provides this option and it is used for accessing various
primitive JavaScript operations. For example, the following is one way to define a function that
converts a value to a number using JavaScript semantics:
%
\begin{equation*}
\begin{array}{l}
 \lbrack\langl\ident{JSEmit}(\str{return \{0\}*1.0;})\rangl\rbrack \\
 \kvd{let}~\ident{number}~(\ignp:\ident{obj}) : \ident{float} ~=~\ident{failwith}~\str{JavaScript stub should not be called.}
\end{array}
\end{equation*}
%
The example uses .NET attributes (on the first line), which are meta-data attached to a function.
When FunScript finds a call to such annotated function, it replaces the call with the specified
JavaScript. The string \textcolor{strclr}{\ttfamily \{0\}} is a placeholder for the argument. Note
that the body of the function is never actually executed and so we use a placeholder.

The approach provides no guarantees about the inline JavaScript code. At compile-time, it cannot
even check that the JavaScript code is syntactically correct. This could be improved using a
parameterized type provider such as $\ident{JSCode}\langl\str{return \{0\}*1.0;}\rangl$, which
would check that the static parameter is syntactically correct JavaScript code. So, type providers
could be used to make this approach safer.

\vspace{-1em}
\paragraph{Other ways of writing annotations.}
Many compilers to JavaScript provide a way for writing type signatures for JavaScript
libraries in the host language. For example, the following snippet shows a declaration of the
jQuery \ident{attr} method from a mapping for js\_of\_ocaml \cite{js_ocmal} (definitions in
SMLtoJs \cite{js_sml} are similar, but simpler):
%
\begin{equation*}
\begin{array}{l}
 \kvd{class type}~\ident{jQuery}~=~\kvd{object} \\
 \quad \kvd{method}~\ident{attr}~:~\ident{js\_string t} ~\rightarrow~ \ident{js\_string t optdef meth} \\
 \quad \kvd{method}~\ident{attr\_set}~:~\ident{js\_string t} ~\rightarrow~ \ident{js\_string t} ~\rightarrow~ \ident{jQuery t meth} \\
 \kvd{end}
\end{array}
\end{equation*}
%
The js\_of\_ocaml project uses the OCaml object model for calling JavaScript libraries.
Compared with FunScript, it is more explicit in importing JavaScript -- the \ident{meth}
type denotes a JavaScript method; \ident{optdef} specifies that the result may be
undefined and \ident{js\_string} denotes a JavaScript string (which is distinct from OCaml
strings). Also, note that the second method is called \ident{attr\_set}. This is a simple
naming trick -- OCaml does not support overloading and js\_of\_ocaml simply ignores anything
after underscore.

Using a type provider is similar to writing a code generator that turns TypeScript
\textcolor{strclr}{\ttfamily d.ts} files into the above OCaml annotation. The main difference
is that type providers provide a smoother integration. Type providers do not produce any artefacts
(generated files) and can import entire repositories lazily.

>>>>

\author{Tomas Petricek
\institute{University of Cambridge, UK}
\email{tomas@tomasp.net}
\and
Don Syme
\institute{Microsoft Research Cambridge, UK}
\email{don.syme@microsoft.com}
\and
Zach Bray
\institute{Type Inferred Ltd}
\email{zachbray@gmail.com}
}
\def\titlerunning{In the Age of Web: Typed Functional-First Programming Revisited}
\def\authorrunning{T. Petricek, D. Syme, Z. Bray}
\begin{document}
\maketitle

@@

Perhaps the most remarkable assumption is the idea that programs fundamentally operate in a closed
world. Although we have learned how to perform FFI and I/O \cite{haskell-ffi,haskell-imperative},
those are treated as dealing with the ``dirty real world''. For a practical solution, we argue that
we need to go much further -- to the extent that deeper integration with (untyped) JavaScript
libraries and (evolving) services inevitably breaks some of the strict type safety requirements.

@@

Note that we use a standard F\# function \ident{List.map} to iterate over the countries. The
function passed as an argument has a side-effect of creating the HTML elements, but it also returns
a new list. The result is a list of $\ident{string}\ast\ident{Indicators}\ast\ident{jQuery}$ values
representing the country name, its indicators (for accessing the World Bank data) and the created
DOM object representing the checkbox.

@@

The aim of this paper is not to make value judgements and argue what is better. Using the case
study as a basis, we claim that the outlined approach is just \emph{one possible} and that it
\emph{works in practice}. The rest of the paper gives more details about the most important
aspects of the approach and discuss alternatives.

@@

Type providers are libraries that are loaded by the compiler (and editor) and are executed at
compile-time (or edit-time). A type provider builds information about types and makes those
available to the compiler. This is done lazily, so the type provider does not need to provide
types for the entire information space at once. In terms of programming language theory, type
providers change the starting point for a type-checking of a program as follows:
%
\begin{equation*}
\hspace{8em}
\begin{array}{rclll}
 \emptyset &\narrow{\vdash}& e : \tau
    &\qquad& (\textit{classical, or closed-world})\\
 \pi~(\includegraphics[width=1em,trim=0 1.5em 0 0]{earth.png})  &\narrow{\vdash}& e : \tau
    &\qquad& (\textit{type providers, or open-world})
\end{array}
\end{equation*}
%
Rather than starting the type-checking with an empty context, we start type-checking with a
context containing a \emph{projection} of some information from the world. This analogy is useful
for understanding how type providers work and how they can go wrong:

\begin{itemize}
\item \emph{Type provider failure.}
  The projection (a type provider) is implemented in F\# and can fail (for example, if the internet
  is required but unavailable). However, when the type provider succeeds, it generates valid F\#
  code and so unchecked compilation errors cannot happen.

@@

\item \emph{Recompilation failure.}
  When the world changes and we \emph{recompile} the program, the result of the projection can
  differ and so we may not be able to type-check code that used to type-check. This can be seen as
  negative aspect, but there is a clear benefit -- it means we discover an error that would happen
  at runtime earlier during recompilation.
\end{itemize}

\noindent
The above points are \emph{general} points about type providers. In this paper, we look at two
\emph{specific} type providers for World Bank and TypeScript. When discussing type providers, we
look at how the provider (or the projection $\pi$) works and what assumptions it makes
about the world. That is, what world changes cause the provided type to change and what world
changes cause a runtime exception.

@@

The World Bank provider \cite{fsharp-data} is a type provider designed specifically for a single
data source. This contrasts with F\# Data type providers for working with CSV, XML and JSON data
that take a sample or schema as a parameter and can thus be used with files of any structure.

The projection implemented by the type provider generates a type called \ident{Countries} that contains
countries as members. Each member returns a value of type \ident{Indicators} that is
also generated and contains all indicators as members. The type provider provides types that are
\emph{erased} during compilation and replaced with runtime implementation. For the World Bank type
provider, the erasure works as follows:
%
\begin{equation*}
\begin{array}{lcl}
 \sem{~\ident{data.}\lident{Czech Republic}~} &=&
   \ident{data.GetCountries}().\ident{GetCountry}(\str{CZE}) \\
 \sem{~\ident{cz.}\lident{School enrollment, tertiary (\% gross)}~} &=&
   \ident{cz.AsyncGetIndicator}(\str{SE.TER.ENRR})
\end{array}
\end{equation*}
%
The underlying operations (\ident{GetCountry}, \ident{GetCountries} and \ident{AsyncGetIndicator})
are operations of an underlying runtime library (which is a normal non-provided library). This is
a typical pattern -- type providers generally produce a light layer on top of a rich runtime library.
An important point here is that the type provider uses the \emph{name} of a country or an indicator
for the member name, but at runtime, it uses a \emph{code} of the country or the indicator. This has
interesting implications for the safety properties.

@@

Given the mapping, we can now look at the assumptions that the World Bank type provider makes about
the world and in which cases it encounters one of the failures defined in Section~\ref{sec:tp-def}.

\begin{itemize}
\item If you are offline when using the type provider (compiling or editing), the type provider will
  not be able to obtain the list of countries or indicators. This is a \emph{type provider failure}.
  The World Bank provider caches the schema, but it still requires internet access for the first time.

\item When a country or indicator is renamed, the compiled code will continue to work (there is no
  \emph{runtime failure}), because the lookup uses the country or the indicator \emph{code}. However,
  recompilation will fail as the member name (which appears in the source code) will be different.

\item Finally, if a country or indicator disappears, we will get \emph{runtime failure} when running
  existing compiled code and \emph{recompilation failure}. Here, the \emph{recompilation
  failure} is a useful indicator that the assumptions made by our code have been violated by
  the data source.
\end{itemize}

\noindent
When contrasted with the traditional view of ML type-safety, the World Bank type provider relaxes
the usual safety conditions. However, this is not the case when compared with standard developer
practices. In particular, if we wrote $\ident{data.GetCountries}().\ident{GetCountry}(\str{CZE})$,
the \emph{runtime failure} behaviour would be the same, but we would lose the useful
\emph{recompilation failure}.

@@

There are two main alternatives worth discussing. First, how would the code look if
we did not use any type provider. Second, what are alternative designs for the World Bank type
provider and how would such alternatives change the safety properties.

\vspace{-1em}
\paragraph{Accessing countries without type providers.} Assuming we want the countries
and regions in Figure~\ref{fig:wb}, we could build a list of country codes and use projection:
%
\begin{equation*}
\begin{array}{l}
 \kvd{let}~\ident{data}~=~\ident{WorldBank.GetDataContext}() \\
 \kvd{let}~\ident{countryCodes}~=~\lbrack~\str{EU},~\str{CZE},~\str{GBP},~\str{USA}~\rbrack \\
 \kvd{let}~\ident{countries}~=~\ident{countryCodes}~|\rangl~\ident{List.map}~(\ident{data.GetCountries}().\ident{GetCountry}())
\end{array}
\end{equation*}
%
The motivation for the case study was a data journalism application with a transparent logic
that is easy to understand and modify. The type provider achieves these goals better -- a country
can be added just by looking through an auto-complete list without knowing its code.
As for the safety properties, using explicit country codes has the same runtime properties as using
the type provider. This could be avoided by listing all countries in a given region, but that is
solving a different problem.

In summary, the World Bank type provider fits a specific niche that we were exploring in the case
study. That is, when we want to create an information analysis that accesses specific data from
an information source. With rising popularity of data journalism and Open Government Data
initiatives, we argue that this is an important problem domain.

@@

However, that is only shifting the burden of error handling from the library to the user. If the
author of the application does not intend to handle errors (for example, by skipping missing data),
they would have to add code that explicitly throws an exception. For the purpose of our case study,
we prefer to keep the end-user code simple -- we can still handle the error using
$\kvd{try}~\ldots~\kvd{with}$.

Whether using option types is more desirable depends on the use case (how we want to handle errors)
and reliability of the data source (countries do not disappear often). The choice can be left to
the user by adding a static parameter (like $\ident{Asynchronous}=\kvd{true}$). For example, the CSV
type provider from F\# Data has a parameter \ident{AssumeMissingValues} that instructs the type
provider to always provide options.

\vspace{-1em}
\paragraph{Accessing data ``as of time''.}
We discussed how to mitigate the problems caused by the fact that the open world changes.
Can we make the world \emph{not} change? This is a sensible question for some data sources
(but not the World Bank). Freebase \cite{google-freebase} supports the notion of ``as of time''.
When calling the API with the \ident{as\_of\_time} parameter it returns data and meta-data that
was present at the specified date.

This would be a desirable option for our case study -- if we want to illustrate point made
by an article, the accompanying application could use data available at the time when the
article was written. However, this feature relies on the ability of the data source -- not all
services have this option.

@@

The projection implemented by the TypeScript type provider imports type annotations for JavaScript
libraries sa specified by TypeScript \cite{ms-typescript}. The following is an excerpt from the
\str{jquery.d.ts} file that describes the \ident{jQuery} value and the \ident{attr} method used
in our example:
%
\begin{equation*}
\begin{array}{l}
 \kvd{declare var}~\ident{jQuery}~:~\ident{JQueryStatic};
\\[0.5em]
 \kvd{interface}~\ident{JQueryStatic}~\{ \\
 \quad (\ident{selector}: \kvd{string}, \ident{context?}: \kvd{any}):~\ident{JQuery}; \\
\}
\\
\kvd{interface}~\ident{JQuery}~\{ \\
\quad \ident{attr}(\ident{attributeName}~:~\kvd{string}):~\kvd{string}; \\
\quad \ident{attr}(\ident{attributeName}~:~\kvd{string},~\ident{value}~:~\kvd{any}):~\ident{JQuery}; \\
\}
\end{array}
\end{equation*}
%
The example demonstrates typical problems that arise when using type providers for language
interoperation. Not all TypeScript language constructs have direct equivalent in F\# and
the TypeScript type provider needs to map them to other constructs that are available:
%
\begin{itemize}
\item The interface file defines \ident{jQuery} as a global variable. F\# type providers cannot
  provide global bindings and so the variable is exposed as a static member of the imported type
  (\emph{i.e.}~\ident{j.jQuery}).

@@

\noindent
It is also worth noting that the \ident{attr} member is overloaded and uses an optional parameter.
However, both of these features are available in F\# and can be directly mapped. The following
snippet shows the runtime code that is generated when the types are erased:

\begin{equation*}
\begin{array}{l}
 \sem{~\ident{j.jQuery.Invoke}(\ident{command})~} ~= \\
 \qquad \kvd{let}~\ident{jq}~=~\ident{Emit.PropertyGetImpl}(\kvd{true}, \str{jQuery}, \lbrack|~|\rbrack) \\
 \qquad \ident{Emit.CallImpl}(\kvd{false}, \str{}, \lbrack|~ \ident{jq};~ \ident{command}~|\rbrack~) \\[0.75em]
 \sem{~\ident{jQuery}(\str{<input>})\ident{.attr}(\str{type},\str{checkbox})~} ~= \\
 \qquad \ident{Emit.CallImpl}(\kvd{false}, \str{attr}, \lbrack|~ \ident{jQuery}(\str{<input>});~\str{type};~\str{checkbox}~|\rbrack~)
\end{array}
\end{equation*}
%
The provided code is never actually \emph{executed}. As discussed in Section~\ref{sec:js-meta}, it
is translated to JavaScript (which then runs in the web browser). So, the above code can be seen
more as instructions for the translator. The mapping is straightforward -- member calls are
translated to \ident{Emit.CallImpl} (empty name denotes that the object itself is called) and
property getters are translated to \ident{Emit.PropertyGetImpl}. The first argument
denotes whether the call is static and the last argument is an array of arguments.

\subsubsection{Safety of cross-language type providers}
TypeScript does not have the traditional type soundness property \cite{ms-safets},
because of covariant generics. However, this is a property of \emph{running} TypeScript code, while
we use TypeScript only as a source of annotations for JavaScript libraries. Still, there is only a
weak guarantee that the library will adhere to the specification. We return to this issue
when discussing translation to JavaScript (Section~\ref{sec:js-meta}).

More importantly, we do not import the typing rules associated with TypeScript.
The type provider mechanism can only provide F\# types that then behave according to the
F\# typing rules. This means that we can import generics\footnote{Due to technical
limitations, this is not currently allowed by F\# type providers, but there are no
fundamental reasons for this.}, but this will not automatically allow using them
covariantly. In some cases, the F\# type system is simply more strict -- and in that case,
users have to use unsafe operations, such as $\ident{unbox}\langl \ident{bool} \rangl$
in the \ident{render} function. We return to the topic of unsafe operations in
Section~\ref{sec:js-meta}.

The above discussion highlights a broader point about using type providers for language
interoperability. For any language, its type system may be weaker or stricter in some ways:

\begin{itemize}
\item When interoperating with a weaker system, the type provider may need
  to map more types to a general type like \ident{object}. This makes the provided operations hard
  to use (as the source language is more flexible than F\#). An alternative is to use explicit
  annotations (like \textcolor{strclr}{\ttfamily d.ts} files for JavaScript).

\item If the imported language has a more precise type system than F\#, the type
  provider has to drop some of the information. This can be done safely for types in
  contravariant positions. In the unsafe case, the runtime needs to perform dynamic checks
  (Section~\ref{sec:conc-interop}).
\end{itemize}

\noindent
In case of TypeScript, we encounter both cases. In many cases, the return type is not
statically known and is exposed as \kvd{any}, which we then map to F\# \ident{object}
(and the developer has to use \ident{unbox}). However, TypeScript also supports limited
form of dependent typing (overloading on constants \cite{ms-typescript-09}). This cannot
be expressed in the F\# type system and the developer has to choose the right provided overload.

@@

\paragraph{Embedding inline JavaScript.}
In general, the host language should not know about the sytnax of the embedded language. This can
be hanlded via quasi-quotations \cite{haskell-templ}, or by embedding JavaScript as strings (perhaps
with syntax checking at a later stage). In FunScript, the \ident{number} function is defined as:
%
\begin{equation*}
\begin{array}{l}
 \lbrack\langl\ident{JSEmit}(\str{return \{0\}*1.0;})\rangl\rbrack \\
 \kvd{let}~\ident{number}~(\ignp:\ident{obj}) : \ident{float} ~=~\ident{failwith}~\str{JavaScript stub should not be called.}
\end{array}
\end{equation*}
%
The example uses .NET attributes, which are meta-data attached to a function. When FunScript finds
a call to such annotated function, it replaces the call with the specified JavaScript.
The approach provides no guarantees about the inline JavaScript code. In practice, this feature
is used only for minimal JavaScript runtime, but the approach could be improved using a
parameterized type provider such as $\ident{JSCode}\langl\str{return \{0\}*1.0;}\rangl$, which
would check that the static parameter is syntactically correct JavaScript code. So, type providers
could be used to make this approach safer.

\vspace{-1em}
\paragraph{Other ways of writing annotations.}
Many compilers to JavaScript provide a way for writing type signatures for JavaScript
libraries in the host language. For example, the following shows a declaration of the
jQuery \ident{attr} method from a mapping for js\_of\_ocaml \cite{js_ocmal} (definitions in
SMLtoJs \cite{js_sml} are simpler):
%
\begin{equation*}
\begin{array}{l}
 \kvd{class type}~\ident{jQuery}~=~\kvd{object} \\
 \quad \kvd{method}~\ident{attr}~:~\ident{js\_string t} ~\rightarrow~ \ident{js\_string t optdef meth} \\
 \quad \kvd{method}~\ident{attr\_set}~:~\ident{js\_string t} ~\rightarrow~ \ident{js\_string t} ~\rightarrow~ \ident{jQuery t meth} \\
 \kvd{end}
\end{array}
\end{equation*}
%
The js\_of\_ocaml project uses the OCaml object model for calling JavaScript libraries.
Compared with FunScript, it is more explicit in importing JavaScript -- the \ident{meth}
type denotes a JavaScript method; \ident{optdef} specifies that the result may be
undefined and \ident{js\_string} denotes a JavaScript string. Also, note that the second method
is called \ident{attr\_set}. This is a simple naming trick -- OCaml does not support overloading
and js\_of\_ocaml simply ignores anything after underscore.

Using a type provider is similar to writing a code generator that turns TypeScript
\textcolor{strclr}{\ttfamily d.ts} files into the above OCaml annotation. The main difference
is that type providers do not produce any artefacts
(generated files) and can import entire repositories lazily.

A wide range of choices is available when targeting JavaScript. The compilation to JavaScript can be
implemented as compiler back-end or as a library (Section~\ref{sec:js-meta}) and there are different
approaches to base libraries (Section~\ref{sec:js-lib}) and asynchronous computing
(Section~\ref{sec:js-async}).

The FunScript library is based on F\# quotations \cite{fsharp-metaprog}. This means that the code
is compiled as ordinary F\#, but the compiler also stores marked blocks of code as data.
In FunScript, we typically need to translate the entire source file. To allow this, we instruct the
compiler to store code as data for the whole module using the \ident{ReflectedDefinition} attribute.
This is done at the beginning of the file:
